Equalizer Programming Guide

### 2008-03-10 Developer/ProgrammingGuide/paper.tex
Update Programming Guide and Quickstart presentation, minor code cleanup
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[x] Cleanup
[ ] Optimization
[x] Documentation

\date{
  \vspace{2cm}
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vspace{1cm}
%  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
  Version 1.1.11 for Equalizer 0.4+svn\\[\medskipamount]
  \today
}

@@

\thispagestyle{empty}
\section*{Equalizer 0.4 Programming Guide}
\subsection*{Contributors}

@@

\clearpage
\thispagestyle{empty}
\tableofcontents
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.0.1   & Nov 06, 2007 & Fixed Typo\\
    1.1     & Nov 12, 2007 & Adapted distributed objects\\
    1.1.1   & Nov 16, 2007 & Minor improvements based on lulu.com print\\
    1.1.2   & Nov 19, 2007 & Minor improvements based on feedback\\
    1.1.3   & Nov 23, 2007 & Add OpenGL extension handling\\
    1.1.4   & Nov 26, 2007 & Changes caused by removal of glX event thread\\
    1.1.5   & Dec 10, 2007 & NodeFactory API changes\\
    1.1.6   & Dec 17, 2007 & Add Scalable Rendering chapter\\
    1.1.7   & Dec 20, 2007 & Per-window object manager changes\\
    1.1.8   & Jan 15, 2008 & Detailed window initialization based on
                             latest refactoring\\
    1.1.9   & Jan 16, 2008 & Detail distributed object usage\\
    1.1.10  & Jan 28, 2008 & Add creation thread for NodeFactory entities\\
    1.1.11  & Feb 20, 2008 & Add advanced chapter for thread synchronization\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

Equalizer provides a framework for the development of parallel OpenGL
applications. Equalizer-based applications can run from a single
shared-memory system with one or multiple graphics cards up to
large-scale graphics clusters. This Programming Guide introduces the
programming interface, often using the Equalizer \textsf{eqPly} example
as a guideline.

@@

Compiling Equalizer is as simple as running \textsf{make} on Linux or
building the Equalizer Visual Studio 2005 solution on Windows. On Mac OS
X 10.4 (Tiger), some prerequisites have to be installed before running
\textsf{make}, as explained in \textsf{README.Darwin}. Mac OS X 10.5
Leopard does have all the prerequisites installed by default.

@@

Any parallel rendering application keeps this basic execution model and
extends it by separating the rendering code from the main event
loop. This rendering coded is then executed in parallel on different
resources. This model is followed by Equalizer, thus making application
development as easy as possible.

@@

The \textsf{eqHello} redraw function renders six colored quads, rotating
around the origin. The \textsf{frameDraw} meth\-od provided by the
\textsf{eq::Channel} can be used as a convience function to setup the
frustum and other OpenGL state. After setting up some additional
lighting parameter, \textsf{eqHello} rotates the scene and renders the
quads using immediate mode:

@@

Choosing the right mode for the application profile is critical for
performance. Equalizer uses the concept of compounds to describe the
task decomposition and result recomposition. It allows the combination
of the different compound modes in any possible way, which allows the
address different bottlenecks in a flexible way.

@@

The advantage of this mode is a low, constant IO overhead for the pixel
transfers, since only color information has to be transmitted. The upper
limit is the number of pixels of the destination view.

It's disadvantage is that it relies on view frustum culling to reduce
the amount of data rendered. Depending on the application data
structure, the overlap of some primitives between individual tiles
limits the scalability of this mode, typically to around eight graphics
cards. Each node has to potentially hold the full database for rendering.

@@

This mode provides very good scalability, since each rendering unit
processes a part of the database. This allows to lower the requirements
on all parts of the rendering pipeline: main memory usage, IO bandwidth,
GPU memory usage, vertex processing and fill rate.

@@

The application has to partion the database so that the rendering units
render only part of the database. Some OpenGL features, do not work
correctly (anti-aliasing) or need special attention (transparency).

The best use of database compounds is to divide the data to a managable
size, and then to use other decomposition modes to achieve more
scalability. Volume rendering is one of the applications which can
profit from database compounds.

@@

Pixel compounds only work very well for purely fill-limited
applications, techniques like frustum culling do not reduce the rendered
data for the source rendering resources.

@@

This per-node frame synchronization does not inhibit latency across
rendering nodes. Furthermore, advanced rendering software which
multi-buffers the dynamic parts of the database can disable the per-node
frame synchronization, as explained in \sref{sNodeFrame}. Some scene
graphs for example do implement multi-buffered data.

@@

For testing and development purposes it is possible to use multiple
instances for one resource, e.g. to run multiple render client nodes on
one computer. For deployment, one node and pipe should be used for each
computer and graphics card, respectively.

@@

The \textsf{eq::Window} class holds a drawable and an OpenGL
context. The drawable can be an on-screen window or an
off-screen\footnote{Off-screen drawables are not implemented yet, but
  can be created by the application and used with Equalizer} PBuffer or
framebuffer object (FBO). The window holds window-system-specific
handles to the drawable and context, e.g., an X11 window \textsf{XID} and
\textsf{GLXContext} for the glX window system.

@@

The \textsf{eq::Channel} class is the abstraction of an OpenGL viewport
within its parent window. It is the entity executing the actual
rendering. The channel's viewport might be overwritten when it is
rendering for another channel during scalable rendering.

@@

\subsubsection{Compound Classification}
The channels of the leaf compounds in the compound tree are designated
as source channels. The top-most channel in the tree is the destination
channel. Only source channels execute rendering task. All channels in a
compound tree work for the destination channel. The destination channel
defines the 2D pixel viewport rendered by all leaf compounds. The
destination channel and pixel viewport can not be overridden by child
compounds.

\subsubsection{Decomposition - Attributes}
Compounds have attributes which configure the decomposition of the
destination's channel viewport, frustum and database. A
\textsf{viewport} decomposes the destination channel and frustum in
screen space. A \textsf{range} tells the application to render a part of
its database, and an \textsf{eye} rendering pass can selectively render
different stereo passes. Setting one or multiple attributes causes the
parent's view to be decomposed accordingly. Attributes are cumulative,
that is, intermediate compound attributes affect and therefore decompose
the rendering of all their children.

@@

When the configuration was successfully initialized, the main rendering
loop is executed. It runs until the user exits the
configuration, or when a maximum number of frames has been rendered,
specified by a command-line argument. The latter is useful for
benchmarks. The \textsf{Clock} is reused for measuring the overall
performance. A new frame is started using \textsf{Config::startFrame}
and a frame is finished using \textsf{Config::finishFrame}.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \subfigure[]{\label{fSync}
    \includegraphics[width=.25\textwidth]{images/sync}}\hfil
  \subfigure[]{\label{fAsync}
    \includegraphics[width=.25\textwidth]{images/async}}%
  {\caption{\small\label{fSyncAsync}Synchronous and asynchronous execution}}
\end{wrapfigure}
\fig{fSyncAsync} shows the execution of (hypothetical) rendering tasks
without latency (\fig{fSync}) and with a latency of
one frame (\fig{fAsync}). With \textsf{eqPly}, a speedup of 15\% has been
observed on a five-node rendering cluster when using a latency of one frame
instead of no
latency\footnote{\link{http://www.equalizergraphics.com/scalability.html}}. A
latency of one or two frames is normally not perceived by the user.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/configInit.pdf}
  {\caption{\small\label{fConfigInit}Config Initialization Sequence}}
\end{wrapfigure}
The config initialization happens in parallel, that is, all config
initialization tasks are transmitted by the server at once and their
completion is synchronized afterwards.

The tasks are executed by the node and pipe threads in parallel. The
parent's initialization methods are always executed before any child
initialization method. This parallelization is necessary to allow a
speedy startup of the configuration on large-scale graphics clusters. On
the other hand, it means that initialization functions are called even
if the parent's initialization has failed.

@@

This section discusses some additional important features not covered
by the previous \textsf{eqPly} section. Where possible code examples
from the Equalizer distribution are used.

@@

Event handling requires a lot of flexibility. On one hand, the
implementation differs slightly for each operating and window system due
to conceptual differences in the OS-specific implementation. On the
other hand, each application and widget set has its own model on how
events are to be handled. Therefore, event handling in Equalizer is
customizable at any stage of the processing, to the extreme of making it
possible to disable all event-related code in Equalizer. In this aspect,
Equalizer substantially differs from GLUT, which imposes an event model
and hides most of the event handling in \textsf{glutMainLoop}. More
information on event handling can be found on the Equalizer
website\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/eventHandling.html}}.

@@

During window and pipe initialization the event handling is set up. For
both entities, \textsf{initEvent\-Handler} is called to register the
pipe or window with an event handler. This method may be overwritten to
use a custom event handler, or to not install an event handler at all
and so disable event handling. Likewise, \textsf{exitEventHandler} is
called to de-initialize event handling.

@@

Events are received by an event handler. The event handler finds the
\textsf{eq::Window} associated to the event. It then creates a generic
\textsf{WindowEvent}, which holds important event data in a 
window system-independent format. The original event is attached to the
generic window event.

@@

User-defined types start at \textsf{eq::ConfigEvent::USER}, and the
member variable \textsf{ConfigEvent::user} can be used to store
\textsf{EQ\_USER\_EVENT\_SIZE}\footnote{currently 32 bytes} bytes. In
this space, the channel's name is stored. Additional variables are used
to transport the pixel format and type, the size and the time it took
for rendering.

@@

Please note that the default thread synchronization synchronizes all
\textsf{Channel::frameDraw} operations on a \textbf{single} node with
the node's main thread. The per-node frame synchronization does not
break the asynchronous execution across nodes. Advanced applications can
remove the per-node frame synchronization.

@@

The local synchronization consists of \textsf{releaseFrameLocal} to
unlock the local frame, and of \textsf{waitFrameLocal} to wait for the
unlock. For the default synchronization model, Equalizer uses the task
method \textsf{frameDrawFinish} which is called on each resource entity
after the last \textsf{Channel::frameDraw} invocation for this
frame. Consequently, \textsf{Pipe::frameDrawFinish} calls
\textsf{Pipe::releaseFrame\-Lo\-cal} to signal that it is done drawing the
current frame, and \textsf{Node::frameDrawFinish} calls
\textsf{Pipe::waitFrameLocal} for each of its pipes to block the node
thread until the current frame has been drawn.

@@

The second, global synchronization is used for the frame completion
during \textsf{Config::finishFrame}, which causes \textsf{frameFinish}
to be called on all entities, passing the oldest frame number, i.e., frame
\textsf{current-latency}. The \textsf{frameFinish} task methods have to
call \textsf{releaseFrame} to signal that the entitity is done with the
frame. The release causes the parent's \textsf{frameFinish} to be
invoked, which is so far synchronized internally. Once all
\textsf{Node::releaseFrame} have been called,
\textsf{Config::finishFrame} returns.

@@

Equalizer uses GLEW\footnote{\link{http://glew.sourceforge.net}} for
OpenGL extension handling, particularly the GLEW MX implementation for
multi-context support.

@@

Functions called from another place need to have the macro or function
\textsf{glewGetContext} that returns the pointer to the GLEWContext of
the appropriate window.

@@

This section explains window initialization in detail. It discussed in
detail the handling on the different window systems. The entry point for
the window initialization is the task method
\textsf{Window::configInit}. This task method first calls a
window-system-specific task method, and then
\textsf{Window::configInitGL} to do generic OpenGL state setup, as shown
in \fig{fWindowInit}.

@@

The drawable creation method \textsf{configInitAGLDrawable} calls either
\textsf{configInitAGLFullscreen} or \textsf{configInitAGLWindow}.

@@

    const bool success = configInitAGLDrawable();
    if( success && !_xDrawable )
    {
        setErrorMessage( "configInitAGLDrawable did set no AGL drawable" );
        return false;
    }

@@

\begin{enumerate}
\item\textsf{getWGLDeviceContext} is used to get a device context which
  might be needed for window creation. Ordinary windows do not need a
  device context during creation, but PBuffers do need one. If a context
  is returned, a function pointer to delete the returned device context
  is set as well.
\item\textsf{configInitWGLDrawable} creates the drawable. The device
  context passed to \textsf{configInitWGLDrawable} is not used when
  creating a window. On success, it sets the window handle.
\item\textsf{chooseWGLPixelFormat} chooses a pixel format based on the
  window attributes. If no device context is given, it uses the device
  context from this window's window handle. The chosen pixel format is
  set on the window's device context as well as on the passed device
  context. The window needs a window handle set before this method is
  called.
\item\textsf{createWGLContext} creates an OpenGL rendering context using
  the given device context. The device context can not be 0. It does not
  set the window's OpenGL context.
\end{enumerate}

@@

{\footnotesize\begin{lstlisting}
bool Window::configInitWGL()
{
    PFNEQDELETEDCPROC deleteDCProc = 0;
    HDC dc = getWGLDeviceContext( deleteDCProc );
    EQASSERT( !dc || deleteDCProc );

    if( !configInitWGLDrawable( dc ))
    {
        if( dc )
            deleteDCProc( dc );
        return false;
    }

    if( !_wglWindowHandle )
    {
        if( dc )
            deleteDCProc( dc );
        setErrorMessage( "configInitWGLDrawable did set no WGL drawable" );
        return false;
    }

@@

    HDC windowDC = GetDC( _wglWindowHandle );
    HGLRC context = createWGLContext( dc ? dc : windowDC );
    ReleaseDC( _wglWindowHandle, windowDC );

    setWGLContext( context );

@@

In immersive visualization, the observer is tracked in and the view
frustra are adapted to the viewer's position and orientation, as shown
in \fig{fTracked}. The transformation origin $\rightarrow$ viewer is set by
the application using \textsf{Config::setHeadMatrix}, which is used by
the server to compute the frustra. The resulting off-axis frustra are
positioned using the channel's head transformation.

@@

The source channels producing one or more \textsf{outputFrame}s use
\textsf{Channel::frame\-Read\-back} to read the pixel data from the frame
buffer. The channels receiving one or multiple \textsf{inputFrame}s use
\textsf{Channel::frameAssemb\-le} to assemble the pixel data into the
framebuffer. Equalizer takes care of the network transport of frame
buffer data between nodes, if needed.

@@

In direct send compositing, each rendering channel is also responsible
for the sort-last composition\footnote{depth-based for polygonal data or
  sorted-blend for volume data} of one screen-space tile. He receives
the framebuffer pixels for his tile from all the other channels. The
size of one tile decreases linearly with the number of source channels,
which keeps the total amount of pixel data per channel constant.

After performing the sort-last compositing, the color information is
transferred to the destination channel, similarly to an 2D (sort-first)
compound. The amount of pixel data for this part of the compositing
pipeline also approaches a constant value, i.e., the full frame buffer.

\fig{fDirectSend} illustrates this algorithm for three channels. The
Equalizer website contains a presentation explaining this
algorithm\footnote{\link{http://www.equalizergraphics.com/documents/EGPGV07.pdf}}.

The following operations have to be possible in order to perform this
algorithm:
\begin{itemize}
\item Selection of color and/or depth frame buffer attachments
\item Restricting the read back area to a part of the rendered area
\item Positioning the pixel data correctly on the receiving channels
\end{itemize}

@@

When using 2D (sort-first) or stereo decompositions, no special
programming is needed to achieve good scalability, as \textsf{eVolve} is
mostly fill-limited and therefore scales nicely in this mode.

@@

Conceptually, the individual volume bricks of each of the source
channels produces pixel data which can be handled like one big 'slice'
through the full texture. Therefore they have to be blen\-ded
back-to-front in the same way as the slice planes during rendering.

DB compounds have the advantage of scaling any part of the volume
rendering pipeline: texture and main memory (smaller bricks for each
channel), fill rate (less samples per channel) and IO bandwidth for
time-dependent data (less data per time step and channel). Since the
amount of texture memory needed for each node decreases linearly, they
make it possible to render data sets which are not feasible to
visualize with any other approach.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/volResult.png}
  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
\end{wrapfigure}
For orthographic rendering, determining the compositing order of the
input frames is trivial. The screen-space orientation of the volume
bricks determines the order in which they have to be composited. The
bricks in \textsf{eVolve} are created by slicing the volume along one
dimension. Therefore the range of the resulting frame buffer images,
together with the sorting order, is used to arrange the frames during
compositing. \fig{fBlendOrtho} shows this composition for one view.

@@

In order to compute the assembly order, \textsf{eVolve} uses the angle
between the origin$\rightarrow$slice vector and the near plane, as shown
in \fig{fBlendPerspective}. When the angle becomes greater than
90\textdegree, the compositing order of the remaining frames has to be
changed. The result image of this composition naturally looks the same
as the volume rendering would when rendered on a single
channel. \fig{fVolResult} shows the result of the composition from
\fig{fBlendPerspective}.

>>>>

\date{
  \vspace{2cm}
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vspace{1cm}
%  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
  Version 1.1.12 for Equalizer 0.5-RC\\[\medskipamount]
  \today
}

@@

\thispagestyle{empty}
\section*{Equalizer 0.5 Programming Guide}
\subsection*{Contributors}

@@

\clearpage
\thispagestyle{empty}
\tableofcontents
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.0.1   & Nov 06, 2007 & Fixed Typo\\
    1.1     & Nov 12, 2007 & Adapted distributed objects\\
    1.1.1   & Nov 16, 2007 & Minor improvements based on lulu.com print\\
    1.1.2   & Nov 19, 2007 & Minor improvements based on feedback\\
    1.1.3   & Nov 23, 2007 & Add OpenGL extension handling\\
    1.1.4   & Nov 26, 2007 & Changes caused by removal of glX event thread\\
    1.1.5   & Dec 10, 2007 & NodeFactory API changes\\
    1.1.6   & Dec 17, 2007 & Add Scalable Rendering chapter\\
    1.1.7   & Dec 20, 2007 & Per-window object manager changes\\
    1.1.8   & Jan 15, 2008 & Detailed window initialization based on
                             latest refactoring\\
    1.1.9   & Jan 16, 2008 & Detail distributed object usage\\
    1.1.10  & Jan 28, 2008 & Add creation thread for NodeFactory entities\\
    1.1.11  & Feb 20, 2008 & Add advanced chapter for thread synchronization\\
    1.1.12  & Mar 10, 2008 & General review for 0.5 release\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

Equalizer provides a framework for the development of parallel OpenGL
applications. Equalizer-based applications can run from a single
shared-memory system with one or multiple graphics cards up to
large-scale graphics clusters. This Programming Guide introduces the
programming interface, often using the Equalizer example \textsf{eqPly}
as a guideline.

@@

Compiling Equalizer is as simple as running \textsf{make} on Linux and
Mac OS X or building the Equalizer Visual Studio 2005 solution on
Windows. Note that on Mac OS X 10.4 (Tiger), some prerequisites have to
be installed before running \textsf{make}, as explained in
\textsf{README.Darwin}.

@@

A parallel rendering application uses the same basic execution model and
extends it by separating the rendering code from the main event
loop. The rendering code is then executed in parallel on different
resources. This model is naturally followed by Equalizer, thus making
application development as easy as possible.

@@

The \textsf{eqHello} redraw function renders six colored quads, rotating
around the origin. The \textsf{frameDraw} meth\-od provided by the
\textsf{eq::Channel} can be used as a convience function to setup the
frustum and other OpenGL state. After setting up some lighting
parameters, \textsf{eqHello} rotates the scene and renders the quads
using immediate mode:

@@

Choosing the right mode for the application profile is critical for
performance. Equalizer uses the concept of compounds to describe the
task decomposition and result recomposition. It allows the combination
of the different compound modes in any possible way, which allows to
address different bottlenecks in a flexible way.

@@

The advantage of this mode is a low, constant IO overhead for the pixel
transfers, since only color information has to be transmitted. The upper
limit is the amount of pixel data for the destination view.

Its disadvantage is that it relies on view frustum culling to reduce the
amount of data submitted for rendering. Depending on the application
data structure, the overlap of some primitives between individual tiles
limits the scalability of this mode, typically to around eight graphics
cards. Each node has to potentially hold the full database for
rendering.

@@

This mode provides very good scalability, since each rendering unit
processes only a part of the database. This allows to lower the
requirements on all parts of the rendering pipeline: main memory usage,
IO bandwidth, GPU memory usage, vertex processing and fill rate.

@@

The application has to partion the database so that the rendering units
render only part of the database. Some OpenGL features do not work
correctly (anti-aliasing) or need special attention (transparency).

The best use of database compounds is to divide the data to a managable
size, and then to use other decomposition modes to achieve further
scalability. Volume rendering is one of the applications which can
profit from database compounds.

@@

Pixel compounds work well for purely fill-limited applications,
techniques like frustum culling do not reduce the rendered data for the
source rendering resources.

@@

This per-node frame synchronization does not inhibit latency across
rendering nodes. Furthermore, advanced rendering software which
multi-buffers the dynamic parts of the database can disable the per-node
frame synchronization, as explained in \sref{sNodeFrame}. Some scene
graphs for example do implement multi-buffered data, and can profit from
relaxing the frame synchronization.

@@

For testing and development purposes it is possible to use multiple
instances for one resource, e.g. to run multiple render client nodes on
one computer. For optimal performance during deployment, one node and
pipe should be used for each computer and graphics card, respectively.

@@

The \textsf{eq::Window} class holds a drawable and an OpenGL
context. The drawable can be an on-screen window or an off-screen
PBuffer. Framebuffer object (FBO) support will be implement by later
version of Equalizer. The window holds window-system-specific handles to
the drawable and context, e.g., an X11 window \textsf{XID} and
\textsf{GLXContext} for the glX window system.

@@

The \textsf{eq::Channel} class is the abstraction of an OpenGL viewport
within its parent window. It is the entity executing the actual
rendering. The channel's viewport is overwritten when it is rendering
for another channel during scalable rendering.

@@

\subsubsection{Compound Classification}
The channels of the leaf compounds in the compound tree are designated
as source channels. The top-most channel in the tree is the destination
channel. Only source channels execute rendering tasks. All channels in a
compound tree work for the destination channel. The destination channel
defines the 2D pixel viewport rendered by all leaf compounds. The
destination channel and pixel viewport can not be overridden by child
compounds.

\subsubsection{Decomposition - Attributes}
Compounds have attributes which configure the decomposition of the
destination's channel viewport, frustum and database. A
\textsf{viewport} decomposes the destination channel and frustum in
screen space. A \textsf{range} tells the application to render a part of
its database, and an \textsf{eye} rendering pass can selectively render
different stereo passes. A \textsf{pixel} parameter adapts the frustum
so that the source channel renders every n$th$ line out of m
lines. Setting one or multiple attributes causes the parent's view to be
decomposed accordingly. Attributes are cumulative, that is, intermediate
compound attributes affect and therefore decompose the rendering of all
their children.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \subfigure[]{\label{fSync}
    \includegraphics[width=.25\textwidth]{images/sync}}\hfil
  \subfigure[]{\label{fAsync}
    \includegraphics[width=.25\textwidth]{images/async}}%
  {\caption{\small\label{fSyncAsync}Synchronous and asynchronous execution}}
\end{wrapfigure}
When the configuration was successfully initialized, the main rendering
loop is executed. It runs until the user exits the
configuration, or when a maximum number of frames has been rendered,
specified by a command-line argument. The latter is useful for
benchmarks. The \textsf{Clock} is reused for measuring the overall
performance. A new frame is started using \textsf{Config::startFrame}
and a frame is finished using \textsf{Config::finishFrame}.

@@

\fig{fSyncAsync} shows the execution of (hypothetical) rendering tasks
without latency (\fig{fSync}) and with a latency of
one frame (\fig{fAsync}). With \textsf{eqPly}, a speedup of 15\% has been
observed on a five-node rendering cluster when using a latency of one frame
instead of no
latency\footnote{\link{http://www.equalizergraphics.com/scalability.html}}. A
latency of one or two frames is normally not perceived by the user.

@@

The config initialization happens in parallel, that is, all config
initialization tasks are transmitted by the server at once and their
completion is synchronized afterwards.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/configInit.pdf}
  {\caption{\small\label{fConfigInit}Config Initialization Sequence}}
\end{wrapfigure}
The tasks are executed by the node and pipe threads in parallel. The
parent's initialization methods are always executed before any child
initialization method. This parallelization is necessary to allow a
speedy startup of the configuration on large-scale graphics clusters. On
the other hand, it means that initialization functions are called even
if the parent's initialization has failed.

@@

This section discusses some additional important features not covered
by the previous \textsf{eqPly} section. Where possible, code examples
from the Equalizer distribution are used to illustrate one use case of
the feature.

@@

Event handling requires a lot of flexibility. On one hand, the
implementation differs slightly for each operating and window system due
to conceptual differences in the specific implementation. On the
other hand, each application and widget set has its own model on how
events are to be handled. Therefore, event handling in Equalizer is
customizable at any stage of the processing, to the extreme of making it
possible to disable all event-related code in Equalizer. In this aspect,
Equalizer substantially differs from GLUT, which imposes an event model
and hides most of the event handling in \textsf{glutMainLoop}. More
information on event handling can be found on the Equalizer
website\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/eventHandling.html}}.

@@

During window and pipe initialization the event handling is set up. For
both entities, \textsf{initEvent\-Handler} is called to register the
pipe or window with an event handler. This method may be overwritten to
use a custom event handler, or to not install an event handler at all in
order to disable event handling. Likewise, \textsf{exitEventHandler} is
called to de-initialize event handling.

@@

Events are received by an event handler. The event handler finds the
\textsf{eq::Window} associated to the event. It then creates a generic
\textsf{WindowEvent}, which holds important event data in an independent
format. The original event is attached to the generic window event.

@@

User-defined types start at \textsf{eq::ConfigEvent::USER}, and the
member variable \textsf{ConfigEvent::user} can be used to store up to
\textsf{EQ\_USER\_EVENT\_SIZE}\footnote{currently 32 bytes} bytes. In
this space, the channel's name is stored. Additional variables are used
to transport the pixel format and type, the size and the time it took
for rendering.

@@

Please note that the default thread synchronization synchronizes all
\textsf{Channel::frame\-Draw} operations on a \textbf{single} node with
the node's main thread. The per-node frame synchronization does not
break the asynchronous execution across nodes. Advanced applications can
remove the per-node frame synchronization.

@@

The local synchronization consists of \textsf{releaseFrameLocal} to
unlock the local frame, and of \textsf{waitFrameLocal} to wait for the
unlock. For the default synchronization model, Equalizer uses the task
method \textsf{frameDrawFinish} which is called on each resource
after the last \textsf{Channel::frameDraw} invocation for this
frame. Consequently, \textsf{Pipe::frameDrawFinish} calls
\textsf{Pipe::releaseFrame\-Lo\-cal} to signal that it is done drawing the
current frame, and \textsf{Node::frameDrawFinish} calls
\textsf{Pipe::waitFrameLocal} for each of its pipes to block the node
thread until the current frame has been drawn.

@@

The second, global synchronization is used for the frame completion
during \textsf{Config::finishFrame}, which causes \textsf{frameFinish}
to be called on all entities, passing the oldest frame number, i.e., frame
\textsf{current-latency}. The \textsf{frameFinish} task methods have to
call \textsf{releaseFrame} to signal that the entitity is done with the
frame. The release causes the parent's \textsf{frameFinish} to be
invoked, which is synchronized internally. Once all
\textsf{Node::releaseFrame} have been called,
\textsf{Config::finishFrame} returns.

@@

Equalizer uses GLEW\footnote{\link{http://glew.sourceforge.net}} for
OpenGL extension handling, particularly the GLEW MX implementation
providing multi-context support.

@@

Functions called from another place need to define a macro or function
\textsf{glewGetContext} that returns the pointer to the GLEWContext of
the appropriate window:

@@

This section explains window initialization in detail. It discusses in
detail the handling on the different window systems. The entry point for
the window initialization is the task method
\textsf{Window::configInit}. This task method first calls a
window-system-specific task method, and then
\textsf{Window::configInitGL} to do generic OpenGL state setup, as shown
in \fig{fWindowInit}.

@@

The drawable creation method \textsf{configInitAGLDrawable} calls either
\textsf{configInitAGLFullscreen}, \textsf{configInitAGLWindow} or \textsf{configInitAGLPBuffer}

@@

    return configInitAGLDrawable();
}
\end{lstlisting}}

@@

\begin{enumerate}
\item\textsf{getWGLPipeDC} is used to get an affinity device context,
  which might be needed for window creation. If a context is returned, a
  function pointer to delete the returned device context is set as
  well. This function might return 0, which is not an error.
\item\textsf{chooseWGLPixelFormat} chooses a pixel format based on the
  window attributes. If no device context is given, it uses the system
  device context. The chosen pixel format is set on the passed device
  context.
\item\textsf{configInitWGLDrawable} creates the drawable. The device
  context passed to \textsf{configInitWGLDrawable} is used to query the
  pixel format and is used as the device context for creating a
  PBuffer. If no device context is given, the display device context is
  used. On success, it sets the window handle. Setting a window handle
  also sets the window's device context.
\item\textsf{createWGLContext} creates an OpenGL rendering context using
  the given device context. If no device context is given, the window's
  device context is used. This function does not set the window's OpenGL
  context.
\end{enumerate}

@@

{\footnotesize\begin{lstlisting}
bool Window::configInitWGL()
{
    PFNEQDELETEDCPROC deleteDCProc = 0;
    HDC dc = getWGLPipeDC( deleteDCProc );
    EQASSERT( !dc || deleteDCProc );

@@

    if( !configInitWGLDrawable( dc, pixelFormat ))
    {
        if( dc )
            deleteDCProc( dc );
        return false;
    }

    if( !_wglDC )
    {
        if( dc )
            deleteDCProc( dc );
        setErrorMessage( "configInitWGLDrawable did not set a WGL drawable" );
        return false;
    }

    HGLRC context = createWGLContext( dc );
    setWGLContext( context );

@@

In immersive visualization, the observer is tracked in and the view
frustra are adapted to the viewer's position and orientation, as shown
in \fig{fTracked}. The transformation $origin \rightarrow viewer$ is set by
the application using \textsf{Config::setHeadMatrix}, which is used by
the server to compute the frustra. The resulting off-axis frustra are
positioned using the channel's head transformation, which can be
retrieved using \textsf{Channel::getHeadTransform}.

@@

The source channels producing one or more \textsf{outputFrame}s use
\textsf{Channel::frame\-Read\-back} to read the pixel data from the frame
buffer. The channels receiving one or multiple \textsf{inputFrame}s use
\textsf{Channel::frameAssemb\-le} to assemble the pixel data into the
framebuffer. Equalizer takes care of the network transport of frame
buffer data between nodes.

@@

In direct send compositing, each rendering channel is also responsible
for the sort-last composition of one screen-space tile. He receives
the framebuffer pixels for his tile from all the other channels. The
size of one tile decreases linearly with the number of source channels,
which keeps the total amount of pixel data per channel constant.

After performing the sort-last compositing, the color information is
transferred to the destination channel, similarly to a 2D (sort-first)
compound. The amount of pixel data for this part of the compositing
pipeline also approaches a constant value, i.e., the full frame buffer.

\fig{fDirectSend} illustrates this algorithm for three channels. The
Equalizer website contains a
presentation\footnote{\link{http://www.equalizergraphics.com/documents/EGPGV07.pdf}}
explaining and comparing this algorithm with the binary-swap algorithm.

The following operations have to be possible in order to perform this
algorithm:
\begin{itemize}
\item Selection of color and/or depth frame buffer attachments
\item Restricting the read-back area to a part of the rendered area
\item Positioning the pixel data correctly on the receiving channels
\end{itemize}

@@

When using 2D (sort-first) or stereo decompositions, no special
programming is needed to achieve good scalability, as \textsf{eVolve} is
mostly fill-limited and therefore scales nicely in these modes.

@@

Conceptually, the individual volume bricks of each of the source
channels produces pixel data which can be handled like one big 'slice'
through the full texture. Therefore they have to be blen\-ded
back-to-front in the same way as the slice planes are blended during
rendering.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/volResult.png}
  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
\end{wrapfigure}
DB compounds have the advantage of scaling any part of the volume
rendering pipeline: texture and main memory (smaller bricks for each
channel), fill rate (less samples per channel) and IO bandwidth for
time-dependent data (less data per time step and channel). Since the
amount of texture memory needed for each node decreases linearly, they
make it possible to render data sets which are not feasible to
visualize with any other approach.

@@

For orthographic rendering, determining the compositing order of the
input frames is trivial. The screen-space orientation of the volume
bricks determines the order in which they have to be composited. The
bricks in \textsf{eVolve} are created by slicing the volume along one
dimension. Therefore the range of the resulting frame buffer images,
together with the sorting order, is used to arrange the frames during
compositing. \fig{fBlendOrtho} shows this composition for one view.

@@

In order to compute the assembly order, \textsf{eVolve} uses the angle
between the $origin \rightarrow slice$ vector and the near plane, as shown
in \fig{fBlendPerspective}. When the angle becomes greater than
90\textdegree, the compositing order of the remaining frames has to be
changed. The result image of this composition naturally looks the same
as the volume rendering would when rendered on a single
channel. \fig{fVolResult} shows the result of the composition from
\fig{fBlendPerspective}.
